Large squeezing in multicomponent trapped Bose-Einstein condensates

### 2012-12-12 large_number_squeezing_with_atoms.tex
Yet more paper progress

\begin{abstract}
We examine the feasibility of creating and measuring large relative number squeezing in multicomponent trapped Bose-Einstein condensates.  In the absence of multimode effects, this squeezing can be arbitrarily large, but a range of processes limit the measurable squeezing in realistic trap configurations.  We conclude that high levels of squeezing with large numbers of atoms can only be achieved in one-dimensional traps with very even density profiles.  We introduce a method of maximising the measurable squeezing by using a pi-pulse during the process to improve spatial mode-matching.
\end{abstract}

@@

\section{Model and two mode solutions}
\label{secTwoModeAnalytic}
\begin{itemize}
  \item Lay out the scheme with diagram, define all variable etc
  \item Present multimode analytic solution, details go into an appendix
  \item Plot a few solutions to give an idea of what things look like, plausible times and so on
  \item Maybe point out it agrees with Oberthaler, or save that for discussion on MM effects?
  \item Point out that number *difference* squeezing can often be considerably higher than absolute number squeezing in a single state
\end{itemize}
Our goal is to describe the number squeezing possibilities of a BEC comprising two relevant internal states denoted by $|a\rangle$ and $|b\rangle$.  The Hamiltonian for a pair of coupled atomic internal modes is the spatial integral of the Hamiltonian density, which is given by:
\begin{equation}
\hat{\mathcal{H}} = \sum_{j} \hat{\psi}_j^{\dagger}\left(-\frac{\hbar^2}{2 m}\nabla^2+V_j\right)\hat{\psi}_j 
          + \sum_{i j}\frac{U_{i j}}{2} \hat{\psi}_i^{\dagger} \hat{\psi}_j^{\dagger} \hat{\psi}_j \hat{\psi}_i
          + \kappa \hat{\psi}_a^{\dagger} \hat{\psi}_b + \kappa^* \hat{\psi}_b^{\dagger}  \hat{\psi}_a
\label{eqFieldHamiltonian}
\end{equation}
where $\hat{\psi}_j(\mathbf{x})$ is the annihilation field operator for a particle at position $\mathbf{x}$ and in internal state $|j\rangle$, $m$ is the mass of the atoms, $V_j(\mathbf{x})$ is the trapping potential for atoms in internal state $|j\rangle$, $U_{ij}$ describe the various inter- and intra-mode nonlinearities, and $\kappa$ describes the induced coupling between the two modes.

We begin by considering a simplified version of the problem, assuming that only one of these spatial modes is relevant for each of the condensate's components.  This is done by writing the multimode atomic field operators in terms of a set of basis functions $\hat{\psi}_a({\mathbf{r}},t) = \sum_n u_{n j}({\mathbf{r}}) \hat{a}_{n}(t)$ and $\hat{\psi}_b({\mathbf{r}},t) = \sum_n u_{n}({\mathbf{r}}) \hat{b}_{n}(t)$, where $\hat{a}_{n}$ and $\hat{b}_{n}$ annihilate a particle with normalised spatial mode $u_n$ and internal state $|a\rangle$ or $|b\rangle$ respectively. With this expansion, the first nonlinear term in the multimode Hamiltonian becomes
\begin{equation}
\hat{H} = \frac{U_{aa}}{2} \int dV \sum_{nmpq} u_n^* u_m^* u_p u_q \, \hat{a}^{\dagger}_{n} \hat{a}^{\dagger}_{m} \hat{a}_{p} \hat{a}_{q} 
\end{equation}
When the system remains predominantly in one mode, we can approximate this sum as the single term $\hat{H} = \frac{\chi_{aa}}{2} \hat{a}^{\dagger} \hat{a}^{\dagger} \hat{a} \hat{a}$ where
\begin{equation}
\chi_{aa} = U_{aa} \int |u_i|^4 \, dV.
\label{eqChiUequivalence}
\end{equation}
We proceed similarly for the other terms.  In the parameter regimes we are interested in, the detuning between the two spatial modes will be irrelevant, so setting our zero of energy appropriately, we are left with only a standard Kerr-type nonlinearity arising from atom-atom interactions, as well as a linear coupling between the two modes.  The Hamiltonian governing this reduced system is given by
\begin{equation}
\hat{H} = \frac{\chi_{aa}}{2} \hat{a}^{\dagger} \hat{a}^{\dagger} \hat{a} \hat{a}
          + \chi_{ab} \hat{a}^{\dagger} \hat{a} \hat{b}^{\dagger} \hat{b}
          + \frac{\chi_{bb}}{2} \hat{b}^{\dagger} \hat{b}^{\dagger} \hat{b} \hat{b}
          + \Omega (\hat{a}^{\dagger} \hat{b} + \hat{b}^{\dagger}  \hat{a} )
\label{eqTwoModeHamiltonian}
\end{equation}
where the $\chi_{ij}$ describe the various inter- and intra-mode nonlinearities, and $\Omega$ couples the two modes and allows for population transfer between them.

The system is prepared with all the population in $|a\rangle$, and vacuum in $|b\rangle$. We choose state $|a\rangle$ to be occupied as a coherent state with average number $\langle \hat{a}^{\dagger} \hat{a} \rangle = N$, where $N$ is the total particle number in our system.  As our simulation and experiment are insensitive to the initial phase of the coherent state, this is equivalent to a mixture of coherent states of uncertain phase, which is also equivalent to mixture of Poissonian-distributed number states.  This state is consistent with BEC coherence experiments \cite{Hadzibabic2004}.  
%@article{Hadzibabic2004,
%  title = {Interference of an Array of Independent Bose-Einstein Condensates},
%  author = {Hadzibabic, Zoran and Stock, Sabine and Battelier, Baptiste and Bretin, Vincent and Dalibard, Jean},
%  journal = {Phys. Rev. Lett.},
%  volume = {93},
%  issue = {18},
%  pages = {180403},
%  numpages = {4},
%  year = {2004},
%  month = {Oct},
%  doi = {10.1103/PhysRevLett.93.180403},
%  url = {http://link.aps.org/doi/10.1103/PhysRevLett.93.180403},
%  publisher = {American Physical Society}
%}
At time $t=0$ the coupling $\Omega$ is switched on and applied until time $t=t_1$, at which point it is turned off, resulting in a portion of the population being transferred into mode $|a\rangle$. As the initial state was a coherent state in both modes, after the population transfer both $|a\rangle$ and $|b\rangle$ are described by coherent states, with mean number $\langle \hat{a}^{\dagger} \hat{a} \rangle = n_a$ and $\langle \hat{b}^{\dagger} \hat{b} \rangle = n_b$ respectively.

The coupling $\Omega$ is then switched off until time $t_2$. During this period the atoms interact solely through the nonlinear terms, with no population being transferred. Due to the form of the Hamiltonian, the number variance remains unchanged, but the quadrature variances in the two modes change over time, which can lead to quadrature squeezing \cite{xxx}.

After this hold time $\tau_{\mathrm{hold}} = t_2 - t_1 $, the coupling $\Omega$ is switched back on until time $t_3$, with a phase shift $\phi$ compared to it's first application. During this last stage, the two modes exchange population and the quadrature variance is converted into number variance, in the same way homodyne measurements are used in quantum optics to convert quadrature squeezing into number squeezing, which can be directly measured. In this interpretation, $\phi$ is the relative phase of the strong local oscillator, which allows specific phase angles of quadrature squeezing to be examined.  This experiment can also be interpreted as a Ramsey interferometer with a final beam splitter phase of $\phi$.

@@

\begin{figure}
    \centering
    \includegraphics[width=8cm]{figures/pulse_scheme.eps}
    \caption{Schematic of the pulse sequence and atomic populations as a function of time.  Typically the coupling sequences at the beginning and end of the sequence are much shorter than the hold time, $\tau_{\mathrm{hold}} = t_2 - t_1 $, which is where the majority of the squeezing occurs. \textbf{Unless I just lied}.}
    \label{figPulseScheme}
\end{figure}

This system can be solved analytically, and expressions for quadrature, absolute and relative number squeezing can be derived. We present the solutions here; the full derivation can be found in \ref{appendixTwoModeDerivation}.
If we define
\begin{eqnarray}
s &=& \sin(\Omega (t_3 - t_2)) \label{eqsDef} \\
c &=& \cos(\Omega (t_3 - t_2)) \\
\lambda_{ij} &=& \chi_{ij} \tau_{\mathrm{hold}} / \hbar \\
A &=& \sqrt{n_a} \exp [n_a (e^{-i(\lambda_{aa}-\lambda_{ab})} - 1 )] \\
A_2 &=& n_a \exp [n_a (e^{-2i(\lambda_{aa}-\lambda_{ab})} - 1 )] \\
B &=& -i e^{i \phi}\sqrt{n_b} \exp [n_b (e^{-i(\lambda_{bb}-\lambda_{ab})} - 1 )] \\
B_2 &=& -n_b e^{2i\phi} \exp [n_b (e^{-2i(\lambda_{bb}-\lambda_{ab})} - 1 )] \\
D &=& AB^* - A^*B \label{eqDDef}
\end{eqnarray}
then the expectation values for number, number variance, and number difference variance after appling the final coupling field for a time $\tau = t_3-t_2$ are given by
\begin{eqnarray}
N_a(t_3) &=& n_a c^2 + n_b s^2 + i c s D \\
%
N_b(t_3) &=& n_a s^2 + n_b c^2 - i c s D \\
%
{\mathrm{Var}} [ N_a](t_3) &=& n_a c^2 + n_b s^2 + 2 n_a n_b c^2 s^2 \\
       && + c^2 s^2 (D^2 - e^{i(\lambda_{aa} - \lambda_{bb})} B_2 A_2^* - e^{-i(\lambda_{aa} - \lambda_{bb})} B_2^* A_2) \nonumber \\
       && + i c s D (1-2 n_a c^2 -2 n_b s^2)  \nonumber\\
       && + 2 i c^3 s n_a (e^{-i(\lambda_{aa} - \lambda_{ab})} A B^* - e^{i(\lambda_{aa} - \lambda_{ab})} A^* B ) \nonumber \\
       && + 2 i c s^3 n_b (e^{i(\lambda_{bb} - \lambda_{ab})} A B^* - e^{-i(\lambda_{bb} - \lambda_{ab})} A^* B ) \nonumber\\
%
{\mathrm{Var}} [ N_b](t_3) &=&  n_a s^2 + n_b c^2 + 2 n_a n_b c^2 s^2 \nonumber \\
       && + c^2 s^2 (D^2 - e^{i(\lambda_{aa} - \lambda_{bb})} B_2 A_2^* - e^{-i(\lambda_{aa} - \lambda_{bb})} B_2^* A_2 ) \nonumber \\
       && + i c s D (-1+2 n_a s^2 + 2 n_b c^2)  \nonumber \\
       && + 2 i c^3 s n_b (e^{-i(\lambda_{bb} - \lambda_{ab})} A^* B - e^{i(\lambda_{bb} - \lambda_{ab})} A B^* ) \nonumber \\
       && + 2 i c s^3 n_a (e^{i(\lambda_{aa} - \lambda_{ab})} A^* B - e^{-i(\lambda_{aa} - \lambda_{ab})} A B^* ) \\
%
{\mathrm{Var}} [ N_a - N_b](t_3) &=& n_a + n_b + 4 i c s D (n_a - n_b)(s^2 - c^2) \nonumber \\
       && + 4 c^2 s^2 (2 n_a n_b +D^2 - e^{i(\lambda_{aa} - \lambda_{bb})} A_2^* B_2 - e^{-i(\lambda_{aa} - \lambda_{bb})} A_2 B_2^* ) \nonumber \\
       && + 4 i c s (c^2 - s^2) (n_a e^{-i(\lambda_{aa} - \lambda_{ab})} A B^* - n_a e^{i(\lambda_{aa} - \lambda_{ab})} A^* B 
                     + n_b e^{-i(\lambda_{bb} - \lambda_{ab})} A^* B - n_b e^{i(\lambda_{bb} - \lambda_{ab})} A B^*) \label{eqNumDiffVariance}
\end{eqnarray}
In the case where all the nonlinearities are equal, such that $\chi_{aa} = \chi_{ab} = \chi_{bb}$, the number variances all reduce to those of a coherent state, e.g. ${\mathrm{Var}}[N_a]=N_a$, and no number squeezing is possible.

While there are three independent nonlinearities $\chi_{ij}$ in the Hamiltonian (\ref{eqTwoModeHamiltonian}), the expressions for the number variances only depend on differences.  This means that number squeezing is parameterised by only two independent quantities, for example $\chi_{aa}-\chi_{ab}$ and $\chi_{aa}-\chi_{bb}$.

{\bf{There seems to be this bogus rule of thumb in the literature that the amount of squeezing you can get is related to $\chi_{11}+\chi_{22}-2\chi_{12}$. Is it worth pointing out that this is wrong?  Joe:  Absolutely, but this might be true in some kind of weak squeezing limit, so we should check for some reference to this in the literature.}}
%Can Uaa + Ubb - 2Uab be zero and still get squeezing? Yep, Uaa=0.03, Ubb=0.01, Uab=0.2 gives squeezing for all quantities of interest

While the analytic expressions for the number variances are not particularly edifying, it is easy to evaluate and plot them for any given set of parameters. As an example, in Figure \ref{figTwoModeAnalyticExamples} we plot the solution for the normalized number variance in mode $|a\rangle$ for a particular set of nonlinearities, choosing a number of different values of $\phi$. The plots illustrate the basic features of the problem: During the final coupling period, the normalised number variance undergoes cyclic variation with a period $\tau=0.5 \Omega$, and has a series of minima. It is necessary to choose both the correct phase $\phi$ as well as the correct length of time for the coupling to be applied in order to obtain optimum squeezing.

\begin{figure}
    \centering
    \includegraphics[width=10cm]{figures/analytic_two_mode_examples.eps}
    \caption{Normalized number variance Var$[N_a]/N_a$ in mode $\hat{a}$ as a function of final recombination time plotted on a logarithmic scale, using four different mixing angles $\phi$. Parameters: $n_a = n_b =5 \times 10^5$, $\chi_{aa}=0.04 \hbar, \chi_{ab}=0, \chi_{bb}=0.01\hbar, \tau_{\mathrm{hold}}=4\times 10^{-4}$s. Solid line, $\phi=0.10$; dash-dotted line, $\phi=1.42$; dashed line, $\phi=3.24$; dotted line, $\phi=4.71$. These particular parameters yield 21dB of number squeezing for the $\phi=0.1$ and $\phi=1.42$ cases.} 
    \label{figTwoModeAnalyticExamples}
\end{figure}

\section{Multimode analysis of squeezing} \label{sec:MMdescription}

@@

The first problem is present even if the spatial modes are uncoupled.  The two-mode model shows that the parameters required to obtain best squeezing are dependent on the strength of the nonlinearity.  This nonlinearity is a function of the mode shape and the number of atoms in that mode.  The optimum hold time $\tau_{\mathrm{hold}}$ depends on this product, as does the recombination phase $\phi$.  Different modes in a multimode environment will typically have varying products of nonlinearity and therefore squeeze at different rates, with each mode having a different optimum $\tau_{\mathrm{hold}}$ and optimum recombination phase $\phi$.  This means that the best squeezing will be achieved by choosing values of $\tau_{\mathrm{hold}}$ and $\phi$ that, averaged across the range of densities, result in an overall lowest possible total number variance.

The second problem is due to the number-dependent dynamics of the spatial modes, which is observable even in a semiclassical simulation.  Due to the nonlinear term in the Hamiltonian, the local phase evolution of the atomic field is density dependent.  Consequently the mode shape of the atomic field changes depending on the nonlinearity-density product, which means that unless $U_{aa}|\psi_a({\mathbf{r}})|^2 = U_{bb}|\psi_b({\mathbf{r}})|^2$ for all ${\mathbf{r}}$, the mode shapes will not overlap perfectly during the final coupling pulse. This degraded mode-matching will result in lower efficiency in converting the quadrature squeezing to number squeezing. We show an example of this effect in Figure~\ref{figXXX}.

\textbf{give a plot of the mode in $|a\rangle$ and $|b\rangle$ at the beginning and end of the hold time to show mismatch}

We might speculate that the effects of these first two problems would be lessened by trap geometries that lead to near-constant density profiles for the BEC.  This should enable higher squeezing.  We cannot make strong conclusions without also considering the effects of coupling between spatial modes, however.  The third potential problem is that coupling between modes is inevitable in the presence of a nonlinearity.  Coupling between modes will disturb mode-matching, and mix fields with different phase evolution, which can rapidly destroy squeezing.

An analytic solution to the higher dimensional problem is intractable, so we must use a numerical solution.  We use stochastic methods based on phase space representations to simulate the dynamics of the multimode quantum fields.  These methods alleviate the problem of directly simulating elements of the Hilbert space, the size of which increases exponentially with the number of spatial modes \cite{stochasticRefs}.  Stochastic methods achieve this by finding a sufficiently well-behaved quasi-probability representation for the density matrix, which can then be simulated as the average behaviour of a number of low-dimensional samples.  These sample objects are the same size as the equivalent classical field.  In our case, the only tractable phase space representation is the functional Wigner representation, which is well behaved when the quantum field is approximately Gaussian, as it is for coherent and squeezed states.  Stochastic methods based on the functional Wigner representation have been used to model atom laser behaviour \cite{atomLaserWigner}, example2 \cite{example2Wigner}, and example3 \cite{example3Wigner}.

The equation of motion for the density matrix is defined by the Hamiltonian in Eq.~(\ref{eqFieldHamiltonian}).  This defines the evolution of the functional Wigner distribution of the system, which has a one-to-one correspondence with the density matrix.  The equation of motion for the functional Wigner distribution contains derivatives of third order which we assume to be negligible.  This uncontrolled approximation is called the Truncated Wigner Approximation (TWA), and while it has been used widely on ultracold gases \cite{atomlaserWigner,orOtherUCGWigner}, care must be taken that the simulation remains valid.  \textbf{Here we insert "all the checks I've done to make sure Wigner holds over the parameters we've simulated"}

>>>>

\begin{abstract}
We examine the feasibility of creating and measuring large relative number squeezing in multicomponent trapped Bose-Einstein condensates.  In the absence of multimode effects, this squeezing can be arbitrarily large, but a range of processes limit the measurable squeezing in realistic trap configurations.  We examine these processes, and suggest methods to mitigate them. We conclude that high levels of squeezing with large numbers of atoms is feasible, but can realistically only be achieved in traps with even density profiles in at least two dimensions. We also introduce a method of maximising the measurable squeezing by using a pi-pulse during the process to improve spatial mode-matching.
\end{abstract}

@@

\section{Model and two mode solutions}
\label{secTwoModeAnalytic}
\begin{itemize}
  \item Lay out the scheme with diagram, define all variable etc
  \item Present multimode analytic solution, details go into an appendix
  \item Plot a few solutions to give an idea of what things look like, plausible times and so on
  \item Maybe point out it agrees with Oberthaler, or save that for discussion on MM effects?
  \item Point out that number *difference* squeezing can often be considerably higher than absolute number squeezing in a single state
\end{itemize}
Our goal is to describe the number squeezing possibilities of a BEC comprising two relevant internal states denoted by $|a\rangle$ and $|b\rangle$.  The Hamiltonian for a pair of coupled atomic internal modes is the spatial integral of the Hamiltonian density, which is given by
\begin{equation}
\hat{\mathcal{H}} = \sum_{j} \hat{\psi}_j^{\dagger}\left(-\frac{\hbar^2}{2 m}\nabla^2+V_j\right)\hat{\psi}_j 
          + \sum_{i j}\frac{U_{i j}}{2} \hat{\psi}_i^{\dagger} \hat{\psi}_j^{\dagger} \hat{\psi}_j \hat{\psi}_i
          + \kappa \hat{\psi}_a^{\dagger} \hat{\psi}_b + \kappa^* \hat{\psi}_b^{\dagger}  \hat{\psi}_a
\label{eqFieldHamiltonian}
\end{equation}
where $\hat{\psi}_j(\mathbf{x})$ is the annihilation field operator for a particle at position $\mathbf{x}$ and in internal state $|j\rangle$, $m$ is the mass of the atoms, $V_j(\mathbf{x})$ is the trapping potential for atoms in internal state $|j\rangle$, $U_{ij}$ describe the various inter- and intra-state nonlinearities, and $\kappa$ describes the coupling between the two states.

We begin by considering a simplified version of the problem, assuming that only one of these spatial modes is relevant for each of the condensate's components.  This is done by writing the multimode atomic field operators in terms of a set of basis functions $\hat{\psi}_a({\mathbf{r}},t) = \sum_n u_{n}({\mathbf{r}}) \hat{a}_{n}(t)$ and $\hat{\psi}_b({\mathbf{r}},t) = \sum_n u_{n}({\mathbf{r}}) \hat{b}_{n}(t)$, where $\hat{a}_{n}$ and $\hat{b}_{n}$ annihilate a particle with normalised spatial mode $u_n$ and internal state $|a\rangle$ or $|b\rangle$ respectively. With this expansion, the first nonlinear term in the multimode Hamiltonian becomes
\begin{equation}
\hat{H} = \frac{U_{aa}}{2} \int dV \sum_{nmpq} u_n^* u_m^* u_p u_q \, \hat{a}^{\dagger}_{n} \hat{a}^{\dagger}_{m} \hat{a}_{p} \hat{a}_{q} 
\end{equation}
When the system remains predominantly in one mode, we can approximate this sum as the single term $\hat{H} = \chi_{aa} \hat{a}^{\dagger} \hat{a}^{\dagger} \hat{a} \hat{a}/2$ where
\begin{equation}
\chi_{aa} = U_{aa} \int |u_i|^4 \, dV.
\label{eqChiUequivalence}
\end{equation}
We proceed similarly for the other terms.  In the parameter regimes we are interested in, the detuning between the two spatial modes will be irrelevant, so setting our zero of energy appropriately, we are left with only a standard Kerr-type nonlinearity arising from atom-atom interactions, as well as a linear coupling between the two modes.  The Hamiltonian governing this reduced system is given by
\begin{equation}
\hat{H} = \frac{\chi_{aa}}{2} \hat{a}^{\dagger} \hat{a}^{\dagger} \hat{a} \hat{a}
          + \chi_{ab} \hat{a}^{\dagger} \hat{a} \hat{b}^{\dagger} \hat{b}
          + \frac{\chi_{bb}}{2} \hat{b}^{\dagger} \hat{b}^{\dagger} \hat{b} \hat{b}
          + \Omega (\hat{a}^{\dagger} \hat{b} + \hat{b}^{\dagger}  \hat{a} )
\label{eqTwoModeHamiltonian}
\end{equation}
where the $\chi_{ij}$ describe the various inter- and intra-mode nonlinearities, and $\Omega$ couples the two modes and allows for population transfer between them.

The system is prepared with all the population in $|a\rangle$, and vacuum in $|b\rangle$. We choose state $|a\rangle$ to be occupied as a coherent state with average number $\langle \hat{a}^{\dagger} \hat{a} \rangle = N$, where $N$ is the total particle number in our system.  As our simulation and experiment are insensitive to the initial phase of the coherent state, this is equivalent to a mixture of coherent states of uncertain phase, which is also equivalent to mixture of Poissonian-distributed number states.  This state is consistent with BEC coherence experiments \cite{Hadzibabic2004}.

At time $t=0$ the coupling $\Omega$ is switched on and applied until time $t=t_1$, at which point it is turned off, resulting in a portion of the population being transferred into mode $|b\rangle$. As the initial state was a coherent state in both modes, after the population transfer both $|a\rangle$ and $|b\rangle$ are described by coherent states, with mean number $\langle \hat{a}^{\dagger} \hat{a} \rangle = n_a$ and $\langle \hat{b}^{\dagger} \hat{b} \rangle = n_b$ respectively.

The coupling $\Omega$ is then switched off until time $t_2$. During this period the atoms interact solely through the nonlinear terms, with no population being transferred. Due to the form of the Hamiltonian, the number variance remains unchanged, but the quadrature variances in the two modes change over time, which can lead to quadrature squeezing \cite{johnssonET2007}.

After this hold time $\tau_{\mathrm{hold}} = t_2 - t_1 $, the coupling $\Omega$ is switched back on until time $t_3$, with a phase shift $\phi$ compared to its first application. During this last stage, the two modes exchange population and the quadrature variance is converted into number variance, in the same way homodyne measurements are used in quantum optics to convert quadrature squeezing into number squeezing, which can be directly measured. In this interpretation, $\phi$ is the relative phase of the strong local oscillator, which allows specific phase angles of quadrature squeezing to be examined.  This experiment can also be interpreted as a Ramsey interferometer with a final beam splitter phase of $\phi$.

@@

\begin{figure}
    \centering
    \includegraphics[width=8cm]{figures/pulse_scheme.eps}
    \caption{Schematic of the pulse sequence and atomic populations as a function of time.  Typically the coupling sequences at the beginning and end of the sequence are much shorter than the hold time, $\tau_{\mathrm{hold}} = t_2 - t_1 $, which is where the majority of the squeezing occurs.}
    \label{figPulseScheme}
\end{figure}

This system can be solved analytically, and expressions for quadrature, absolute and relative number squeezing can be derived. We present the solutions here; the full derivation can be found in \ref{appendixTwoModeDerivation}.
If we define
\begin{eqnarray}
s &=& \sin(\Omega (t_3 - t_2)) \label{eqsDef} \\
c &=& \cos(\Omega (t_3 - t_2)) \\
\lambda_{ij} &=& \chi_{ij} \tau_{\mathrm{hold}} / \hbar \\
A &=& \sqrt{n_a} \exp [n_a (e^{-i(\lambda_{aa}-\lambda_{ab})} - 1 )] \\
A_2 &=& n_a \exp [n_a (e^{-2i(\lambda_{aa}-\lambda_{ab})} - 1 )] \\
B &=& -i e^{i \phi}\sqrt{n_b} \exp [n_b (e^{-i(\lambda_{bb}-\lambda_{ab})} - 1 )] \\
B_2 &=& -n_b e^{2i\phi} \exp [n_b (e^{-2i(\lambda_{bb}-\lambda_{ab})} - 1 )] \\
D &=& AB^* - A^*B \label{eqDDef}
\end{eqnarray}
then the expectation values for number, number variance, and number difference variance after appling the final coupling field are given by
\begin{eqnarray}
N_a(t_3) &=& n_a c^2 + n_b s^2 + i c s D \\
%
N_b(t_3) &=& n_a s^2 + n_b c^2 - i c s D \\
%
{\mathrm{Var}} [ N_a](t_3) &=& n_a c^2 + n_b s^2 + 2 n_a n_b c^2 s^2 \\
       && + c^2 s^2 (D^2 - e^{i(\lambda_{aa} - \lambda_{bb})} B_2 A_2^* - e^{-i(\lambda_{aa} - \lambda_{bb})} B_2^* A_2) \nonumber \\
       && + i c s D (1-2 n_a c^2 -2 n_b s^2)  \nonumber\\
       && + 2 i c^3 s n_a (e^{-i(\lambda_{aa} - \lambda_{ab})} A B^* - e^{i(\lambda_{aa} - \lambda_{ab})} A^* B ) \nonumber \\
       && + 2 i c s^3 n_b (e^{i(\lambda_{bb} - \lambda_{ab})} A B^* - e^{-i(\lambda_{bb} - \lambda_{ab})} A^* B ) \nonumber\\
%
{\mathrm{Var}} [ N_b](t_3) &=&  n_a s^2 + n_b c^2 + 2 n_a n_b c^2 s^2 \nonumber \\
       && + c^2 s^2 (D^2 - e^{i(\lambda_{aa} - \lambda_{bb})} B_2 A_2^* - e^{-i(\lambda_{aa} - \lambda_{bb})} B_2^* A_2 ) \nonumber \\
       && + i c s D (-1+2 n_a s^2 + 2 n_b c^2)  \nonumber \\
       && + 2 i c^3 s n_b (e^{-i(\lambda_{bb} - \lambda_{ab})} A^* B - e^{i(\lambda_{bb} - \lambda_{ab})} A B^* ) \nonumber \\
       && + 2 i c s^3 n_a (e^{i(\lambda_{aa} - \lambda_{ab})} A^* B - e^{-i(\lambda_{aa} - \lambda_{ab})} A B^* ) \\
%
{\mathrm{Var}} [ N_a - N_b](t_3) &=& n_a + n_b + 4 i c s D (n_a - n_b)(s^2 - c^2) \nonumber \\
       && + 4 c^2 s^2 (2 n_a n_b +D^2 - e^{i(\lambda_{aa} - \lambda_{bb})} A_2^* B_2 - e^{-i(\lambda_{aa} - \lambda_{bb})} A_2 B_2^* ) \nonumber \\
       && + 4 i c s (c^2 - s^2) (n_a e^{-i(\lambda_{aa} - \lambda_{ab})} A B^* - n_a e^{i(\lambda_{aa} - \lambda_{ab})} A^* B 
                     + n_b e^{-i(\lambda_{bb} - \lambda_{ab})} A^* B - n_b e^{i(\lambda_{bb} - \lambda_{ab})} A B^*) \label{eqNumDiffVariance}
\end{eqnarray}
In the case where all the nonlinearities are equal, such that $\chi_{aa} = \chi_{ab} = \chi_{bb}$, the number variances all reduce to those of a coherent state, e.g. ${\mathrm{Var}}[N_a]=N_a$, and no number squeezing is possible. It should also be noted that, although there are three independent nonlinearities $\chi_{ij}$ in the Hamiltonian (\ref{eqTwoModeHamiltonian}), the expressions for the number variances only depend on differences. This means that number squeezing is parameterised by only two independent quantities, for example $\chi_{aa}-\chi_{ab}$ and $\chi_{aa}-\chi_{bb}$.

%Can Uaa + Ubb - 2Uab be zero and still get squeezing? Yep, Uaa=0.03, Ubb=0.01, Uab=0.02 gives squeezing for all quantities of interest

While the analytic expressions for the number variances are not particularly edifying, it is easy to evaluate and plot them for any given set of parameters. As an example, in Figure \ref{figTwoModeAnalyticExamples} we plot the solution for the normalized number variance in mode $|a\rangle$ for a particular set of nonlinearities, choosing a number of different values of $\phi$. The plots illustrate the basic features of the problem: During the final coupling period, the normalised number variance undergoes cyclic variation with a period $\tau=\pi / \Omega$, and has a series of minima. It is necessary to choose both the correct phase $\phi$ as well as the correct length of time for the coupling to be applied in order to obtain optimum squeezing.

This two-mode analytic model allows arbitrarily good number squeezing, provided there are no limits to the number of particles in the system or to the hold time. Unfortunately, in any realistic system, there are more than two modes present, and multimode effects conspire to limit the level of squeezing that can be achieved. It is to this that we now turn.

\begin{figure}
    \centering
    \includegraphics[width=10cm]{figures/analytic_two_mode_examples.eps}
    \caption{Normalized number variance Var$[N_a]/N_a$ in mode $|a \rangle$ as a function of final recombination time plotted on a logarithmic scale, using four different mixing angles $\phi$. Parameters: $n_a = n_b =5 \times 10^5$, $\chi_{aa}=0.04 \hbar, \chi_{ab}=0, \chi_{bb}=0.01\hbar, \tau_{\mathrm{hold}}=4\times 10^{-4}$s. Solid line, $\phi=0.10$; dash-dotted line, $\phi=1.42$; dashed line, $\phi=3.24$; dotted line, $\phi=4.71$. These particular parameters yield 21dB of number squeezing for the $\phi=0.1$ and $\phi=1.42$ cases.} 
    \label{figTwoModeAnalyticExamples}
\end{figure}

\section{Multimode number squeezing analysis} \label{sec:MMdescription}

@@

The first problem is present even if the spatial modes are uncoupled.  The two-mode model shows that the parameters required to obtain best squeezing are dependent on the strength of the nonlinearity. In the multimode case, the effective nonlinearity of a mode is a function of the mode shape, as is shown by Eq.~(\ref{eqChiUequivalence}). The optimum hold time $\tau_{\mathrm{hold}}$ depends on the product of the effective mode nonlinearity and number number of particles in that mode, as does the recombination phase $\phi$.  Different modes in a multimode environment will typically have varying effective nonlinearity-particle number products and therefore squeeze at different rates, with each mode having a different optimum $\tau_{\mathrm{hold}}$ and optimum recombination phase $\phi$. This means that the best squeezing will be achieved by choosing values of $\tau_{\mathrm{hold}}$ and $\phi$ that, averaged across all the spacial modes present, result in an overall lowest possible total number variance.

The second problem is due to the number-dependent dynamics of the spatial modes, which is observable even in a semiclassical simulation.  Due to the nonlinear term in the Hamiltonian, the local phase evolution of the atomic field is density dependent.  Consequently the mode shape of the atomic field changes depending on the nonlinearity-density product, which means that unless $U_{aa}|\psi_a({\mathbf{r}})|^2 = U_{bb}|\psi_b({\mathbf{r}})|^2$ for all ${\mathbf{r}}$, the mode shapes will not overlap perfectly during the final coupling pulse. This degraded mode-matching will result in lower efficiency when converting the quadrature squeezing to number squeezing.

We might speculate that the effects of these first two problems would be lessened by trap geometries that lead to near-constant density profiles for the BEC.  This should enable higher squeezing.  We cannot make strong conclusions without also considering the effects of coupling between spatial modes, however, which leads to the third potential problem: Coupling between modes is inevitable in the presence of a nonlinearity, meaning that the modes cannot be analyzed independently. Coupling between modes will disturb mode-matching, as well as mixing fields with different phase evolution, which can rapidly destroy squeezing.

An analytic solution to the multimode, higher-dimensional problem is intractable, so we must use a numerical solution.  We use stochastic methods based on phase space representations to simulate the dynamics of the multimode quantum fields.  These methods alleviate the problem of directly simulating elements of the Hilbert space, the size of which increases exponentially with the number of spatial modes \cite{stochasticRefs}.  Stochastic methods achieve this by finding a sufficiently well-behaved quasi-probability representation for the density matrix, which can then be simulated as the average behaviour of a number of low-dimensional samples.  These sample objects are the same size as the equivalent classical field.  In our case, the only tractable phase space representation is the functional Wigner representation, which is well behaved when the quantum field is approximately Gaussian, as it is for coherent and squeezed states.  Stochastic methods based on the functional Wigner representation have been used to model the behaviour or BECs and atom lasers \cite{johnssonET2007,dallET2009,dennisET2010}, as the Wigner representation allows the mapping of the system to a Fokker-Plank equation (unlike the $P$-representation), and does not suffer exponential path-weighting problems that would constrain it to very short simulation intervals (as would the positive-$P$ representation).

The equation of motion for the density matrix is defined by the Hamiltonian in Eq.~(\ref{eqFieldHamiltonian}).  This defines the evolution of the functional Wigner distribution of the system, which has a one-to-one correspondence with the density matrix.  The equation of motion for the functional Wigner distribution contains derivatives of third order which we assume to be negligible.  This uncontrolled approximation is called the Truncated Wigner Approximation (TWA), and while it has been used widely on ultracold gases \cite{atomlaserWigner,orOtherUCGWigner}, care must be taken that the simulation remains valid. For our simulations we checked the results of our simulations against both the analytic two-mode solution, as well as checking for typical indications of TWA breakdown such a the appearance of negative densities in lightly populated modes. In all the regimes we investigated the TWA remained valid, apart from the most demanding simulations which took place in 3D with high nonlinearities and long hold times. The difficulty with these 3D simulations was that due to the large number of modes they included, the number of particles per mode could drop to ten or less in the densest regions where the squeezing was generated. Over very long time scales, this low mode occupation began to result in TWA breakdown. As we could characterise the regime in which breakdown began to occur, however, we were able stay away from it and to ensure our simulations remained valid.
